The Spatial and Temporal Domains of Modern Ecology

### 2017-05-24 paper/SI.tex
Updated analysis of annual change in observing methods and scales. Revision of main text to accomodate this change, as well as addition of related figures to SI.  Fixes on explanation of methods/FAQ in SI.

To record the spatio-temporal dimensions of ecological observations, we considered the observational scales to be those at which a total measurement of the feature(s) of interest was made, rather than the scales that might be represented by the observations.  As an example, if a paper reported data on plant species diversity collected using 10 1 m$^2$ sub-plots randomly placed within a larger plot of 100 m$^2$, of which 10 were placed, we recorded 1 m$^2$ as the spatial resolution and 100 m$^2$ as the total extent of the observation. While the 10 sub-plots might be representative of diversity at larger scales (e.g. the particular plant community), that relationship depends both on sample design (number of replicates, placement, grain/resolution) and the properties of the ecosystem \cite{underwood_experiments_1997,palmer_scale_1994, cao_comparison_2002}. Most studies also did not quantify the broader scales represented by their observations, nor did they report dimensions that would allow these to be independently estimated, such as the maximum distance between sample plots, which could have been used to estimate the spatial extent represented by the sample \cite{palmer_scale_1994}.

Similarly, we did not try to quantify the extent to which values of individual observations (e.g. a single plot) of an ecological feature might be correlated with spatially or temporally adjacent values.  Autocorrelation is common to most ecological phenomena \cite{legendre_spatial_1993,collins_method_2000-1}, and it means that an observation can be representative of the properties within a broader area or time around the point at which it is taken. For example, soil properties are typically measured by analysis of soil cores collected using an augur. The two-dimensional resolution of this measurement is equal to the area of the augur, but the measured soil properties are likely to be very similar, if not identical, to those found in the soil within a radius of at least a few to several hundreds of meters around the core location. However, the distance and strength of this correlation can vary substantially according to the ecological features being observed, and such correlation lengths were reported in almost no studies. Attempting to estimate these lengths, and add them to actual observations scales, would have added further uncertainty to estimates. Moreover, this concern only applies to spatially and temporally discontinuous (or discontiguous) observations, and since many of observations are collected using techniques that permit continuous measurements (e.g. weather data over time or satellite images over space), we elected to use a consistent standard for estimating dimensions.

In assessing spatial scales, our analysis only considered the Cartesian plane. We did not calculate the z, or depth, dimension, although this dimension is of greater importance for certain sub-disciplines of ecology (e.g. depth profiles in marine ecology). In some cases (primarily paleoecological studies), values extracted from the z-dimension provided temporal information that was used to calculate both the interval and the duration of the observation.

@@

A fuller description of the project methodology is contained within the answers to the list of questions below. This set of Frequently Asked Questions (FAQ) was provided to each observer for initial study and reference, in order to ensure methodological consistency (see next section).

@@

   \Hitem \emph{\textbf{What is sampling interval, and how do we record it?}} Sampling interval is the time that elapsed between repeat observations of the same point in space or individual organism. In many cases, observations will only be made one time--list a value of 0 for these.  

@@

   \Hitem \emph{\textbf{What is the effective study duration, and how do estimate it?}} The effective study duration is the total period of time over which the phenomenon of interest was observed. More specifically, in the case of repeated observations, this is the total time that elapsed between when the first and last observations at a given point were made. For once-off (unreplicated) observations, this time is equivalent to the sample duration. However, there may be cases where once-off observations may have a longer effective duration then the sample duration.  For example, consider a study that counts occurrences of pollinators over three years, using transects that are located in different locations within the broader study area during each year \cite{rollin_differences_2013-1}.  The observations are therefore not strictly temporally replicated, but the authors control for year of collection in their subsequent analysis to avoid confounding effects.  In this case, we can consider the effective duration to be three years, as the temporal information is encoded in the analysis. 
  \Hitem \emph{\textbf{What is actual study duration, and how do we calculate it?}} The actual study duration is the integral of sampling duration, or the time spent making one observation of the phenomenon in question. To clarify, duration is the total time spent sampling/observing a single point in space--not the span of time between first and last sample (the effective duration), nor the integral of time spent in observing all spatial replicates. 
  \Hitem \emph{\textbf{Should study duration be the total time spent sampling all sites or the amount of time spent sampling per site (e.g. for 5 minute point counts of birds at 10 sites each repeated twice, should we enter 100 minutes (5 minutes X 2 repeats X 10 sites) or 10 minutes (5 minutes X 2 repeats) for duration)?}} As stated above, duration is the total spent observing a single point in space, so in this case that would be 10 minutes (then converted to days, so 10 / (60 * 24))).
  \Hitem \emph{\textbf{How do we record study duration when there are no repeat observations?}} In these cases, study duration is equal to sampling duration. 
   \Hitem \emph{\textbf{How do I record sampling interval in cases where the interval is inconsistent?  For example, in a study were observations were observations were repeated in 1979, 1980, 1981, 1984, 2007, 2009?}} Find the time between each successive period, and then take the average of that (remember to convert to days!). If there are two or more sets of unevenly spaced days for each site/plot/measurement being taken in the study, then find the average interval for each, and average the averages.  
    \Hitem \emph{\textbf{How do you determine the time between samples for paleo-reconstructions?}} Use the minimum estimate for dating precision as the estimate of time between samples (e.g. 50 years in the study of European charcoal deposits DOI:10.1111/geb.12090). 

@@

    \Hitem \emph{\textbf{What about sampling intervals and durations for instrument-collected observations?}} These are often similar in essence to the paleo-reconstruction case. Take the minimum temperature or daily rainfall recorded at a weather station, which require constant observation over 24 hours to report.  In such cases, the sampling interval and sampling duration are both 24 hours. On the other hand, automated logging systems will provide a series of high frequency observations that are collected instantaneously. In these cases the sampling duration should be $\leq$second, and the sampling interval should be the period between successive instantaneous measurements. 
    \Hitem \emph{\textbf{How do you handle the time between samples for a case where repeated samples are taken during a season, across several seasons (e.g.``we performed repeat bird counts every 10 days between March and June of 2005, 2006, and 2008'')?}} Since the sampling is focused on seasons, and presumably some season-specific phenomenon (e.g. breeding behavior), the reported values should be pegged to the season, not averaged across the full study span (the start and end dates of the study). So in this example, it would be 10 days.

\hspace{-1cm}\underline{\textbf{Spatial scales:}}
  \Hitem \emph{\textbf{What is sampling resolution, and how do I record it?}} This is the finest scale at which a complete measurement of every unit of the quantity of interest is recorded.  For example, if the measurement in question is a tree stem count, the resolution is determined by the size of the plot used to record every tree stem. Taking this example further, let's say a study reports a plot size of 100 x 100 m, but then goes on to report that they counted stems within a single 1 m wide transect within this larger plot.  In this case, the plot resolution is in fact 100 m x 1 m, or 100 m$^2$ (sampling resolution should be reported in m$^2$).  In another example, if the reported plot size was 20 x 20 m, but the authors in fact only measured a random selection of, say, grass stems on which they counted aphids within those plots, then use an estimate of the area of the grass stem as the sampling resolution \cite{gagic_agricultural_2012}.  
  \Hitem \emph{\textbf{What is study extent, and how do I record it?}}  Broadly defined, actual study extent is the sampling resolution multiplied by the number of spatial replicates, divided by 10,000 to convert to hectares. For studies in which the spatial replicates are not spatially contiguous (as with most field-based studies), this means plot resolution (see Q16) multiplied by number of plots. For spatially contiguous studies (e.g., those based on remote sensing imagery), it should be the total area covered by the imagery, i.e., pixel resolution multiplied by the number of pixels. However, only record the area analyzed by the authors. If they used a sub-sample of pixels, the actual extent is the number of those pixels multiplied by pixel resolution.

Effective survey extent is defined as the total extent bounding all spatial replicates, divided by 10,000 to convert to hectares. For studies in which spatial replicates are not spatially contiguous, this means the area of the minimum convex polygon bounding all spatial replicates. For example, if the study is conducted in a national park, the effective survey extent would be the area of the national park; if the study is conducted in three national parks, the effective survey extent would be the sum of the areas of all national parks. To calculate the effective survey extent, use the area of the study area/region given in the paper; when the area is not given, but when the survey region is given by name (e.g., Joshua Tree National Park or United States), look up the area through an online search and convert to hectares. When the area is not named, but a map is given, use an appropriate digitizing platform with a suitable map-providing backend (e.g. Google Earth Pro, QGIS with OpenLayers plugin) to navigate to the region and delineate a minimum convex polygon surrounding the plots/transects/sampling units to calculate the area in hectares. When studies focus on portions of survey regions (e.g., mangroves in a coastal National Park), attempt to delineate the focal portions (the mangroves) and not the larger survey region (entire National Park) using Google Earth Pro.

For spatially contiguous studies (e.g. those based on satellite imagery), the effective survey extent is the total area covered by the imagery (in such cases, effective survey extent equals actual survey extent). However, if spatially contiguous studies only use a sub-sample of pixels, the effective survey extent is the area of the minimum convex polygon bounding all pixels (calculated following the methods above).

@@

@@

  \Hitem \emph{\textbf{What about studies that sample individual organisms?}} If the study is making a total count of all organisms (let's say a mammal species) within a fixed plot size, or even a variable plot size from which an average plot size (and thus sampling resolution) can be estimated, then follow the procedures described in Q16-17. However, if the individual animals are the unit of measure (either because a sub-sample of them is being made within a defined plot, or because the observation is not contingent on being located within a plot (maybe a blood sample or body weight is being recorded, for example), then simply estimate two-dimensional area occupied by the animal, the number of animals sampled, and then calculate area from that. Occasionally individual animals might be recorded, but within the context of some natural feature, such as a nesting site where the survival of individual chicks is the measurement of interest \cite{roche_relative_2008}. In this case, an estimate of the nest area provides the sampling resolution. Another possible case is where individual animals are being tracked using radio or GPS collars. In these cases, use the estimate of the collared/tagged animal's body area, and then treat the number of reported locational observations as the number of spatial replicates, and then use those two values to calculate sampled area. The total GPS points collected from all individuals can be used if we can safely assume that each individual sampled did not cover the same exact location.  If the number of GPS points is not given, the number of fixes can be estimated from the duration during which individuals were collared and the recording interval.

@@

\subsubsection*{Between-observer coefficient of variation}
We used the maximum uncertainty values for each dimension from the inter-observer variability analysis (Table S2) to determine the bounds of the random perturbations applied to each record over the 1000 iteration resample.

%\begin{figure}[!ht]
%%\begin{wrapfigure}{c}{1\textwidth}
%\includegraphics[width=1\textwidth]{figures/res_v_extent_pbstrap2.pdf}
%\vspace{-0.15 cm}
%\caption{Two-dimensional kernel density estimates of observational densities within the domains defined by sampling interval and spatial resolution (left) and temporal duration and spatial extent (right). Density estimates were applied to the log-transformed values of each observational dimension after being randomly perturbed within uncertainty bounds drawn from an assessment of between-observer variability. Density estimates are rescaled to represent percentages. }
%\label{afoto1}
%\end{figure}
%
Figure S1 indicates the effect that varying bandwidth has on the appearance of the kernel density estimates. The smallest bandwidth (0.4) tested (Fig. S2 top) shows that the primary observational concentrations described in the main text are evident but harder to discern.  Most difficult to discern is the oblong concentration of observations that in Fig. 1A (main text) is bounded on the lower right at monthly to yearly intervals and 100-1000 m$^2$ resolutions and on the upper left by near-daily to monthly observations and 0.1-10 ha resolutions. With the smaller bandwidth this concentration appears as two separate patches (Fig. S2 top left), but with 0.7 bandwidth applied becomes coherent (Fig. S3 middle left).

>>>>

We recorded six primary dimensions of ecological observations, three related to space and three related to time.  The space-related dimensions were resolution, extent, and actual extent. Here extent was primarily defined as the area falling within a perimeter defined by the outermost spatial replicates, while actual extent was defined as the summed area of all sample plots (i.e. N * resolution, where N is the number of spatial replicates, which we also recorded), or the area that ecologists observe in practice. In assessing spatial scales, our analysis only considered the Cartesian plane. We did not calculate the z, or depth, dimension, although this dimension is of greater importance for certain sub-disciplines of ecology (e.g. depth profiles in marine ecology). In some cases (primarily paleoecological studies), values extracted from the z-dimension provided temporal information that was used to calculate both the interval and the duration of the observation.

For time dimensions, we extracted information related to interval, duration, and actual duration. Duration was defined as the time between the first and last temporal replicate, whereas actual duration quantifies the amount of time spent observing a particular location, which we calculated by multiplying sampling duration (the time spent collecting a single temporal replicate) by the number of temporal replicates.

A full definition of all dimensions and how they were recorded is contained within the answers to the list of questions below. This set of Frequently Asked Questions (FAQ) was provided to each observer for initial study and reference, in order to ensure methodological consistency (see next section).

%To record the spatio-temporal dimensions of ecological observations, we considered the observational scales to be those at which a total measurement of the feature(s) of interest was made, rather than the scales that might be represented by the observations.  As an example, if a paper reported data on plant species diversity collected using 10 1 m$^2$ sub-plots randomly placed within a larger plot of 100 m$^2$, of which 10 were placed, we recorded 1 m$^2$ as the spatial resolution and 100 m$^2$ as the total extent of the observation.

%While the 10 sub-plots might be representative of diversity at larger scales (e.g. the particular plant community), that relationship depends both on sample design (number of replicates, placement, grain/resolution) and the properties of the ecosystem \cite{underwood_experiments_1997,palmer_scale_1994, cao_comparison_2002}. Most studies also did not quantify the broader scales represented by their observations, nor did they report dimensions that would allow these to be independently estimated, such as the maximum distance between sample plots, which could have been used to estimate the spatial extent represented by the sample \cite{palmer_scale_1994}.

%Similarly, we did not try to quantify the extent to which values of individual observations (e.g. a single plot) of an ecological feature might be correlated with spatially or temporally adjacent values.  Autocorrelation is common to most ecological phenomena \cite{legendre_spatial_1993,collins_method_2000-1}, and it means that an observation can be representative of the properties within a broader area or time around the point at which it is taken. For example, soil properties are typically measured by analysis of soil cores collected using an augur. The two-dimensional resolution of this measurement is equal to the area of the augur, but the measured soil properties are likely to be very similar, if not identical, to those found in the soil within a radius of at least a few to several hundreds of meters around the core location. However, the distance and strength of this correlation can vary substantially according to the ecological features being observed, and such correlation lengths were reported in almost no studies. Attempting to estimate these lengths, and add them to actual observations scales, would have added further uncertainty to estimates. Moreover, this concern only applies to spatially and temporally discontinuous (or discontiguous) observations, and since many of observations are collected using techniques that permit continuous measurements (e.g. weather data over time or satellite images over space), we elected to use a consistent standard for estimating dimensions.

@@

@@

   \Hitem \emph{\textbf{What is interval, and how do we record it?}} Interval is the time that elapsed between repeated observations of the same point in space or individual organism. In many cases, observations will only be made one time--list a value of 0 for these.  

@@

   \Hitem \emph{\textbf{What is the study duration, and how do estimate it?}} The study duration (or, simply, duration) is the total period of time over which the phenomenon of interest was observed. More specifically, in the case of repeated observations, this is the total time that elapsed between the first and last observations at a given point in space (or of the same individual organism or community) were made. For once-off (unreplicated) observations, this time is equivalent to the sample duration. However, there may be cases where once-off observations may have a longer duration then the sample duration.  For example, consider a study that counts occurrences of pollinators over three years, using transects that are located in different locations within the broader study area during each year \cite{rollin_differences_2013-1}. The observations are therefore not strictly temporally replicated, but the authors control for year of collection in their subsequent analysis to avoid confounding effects.  In this case, we can consider the effective duration to be three years, as the temporal information is encoded in the analysis. 
  \Hitem \emph{\textbf{What is \emph{actual duration}, and how do we calculate it?}} The actual duration is the integral of sampling duration, or the time spent making one observation of the phenomenon in question. To clarify, actual duration is the total time spent sampling/observing a single point in space--not the span of time between first and last sample (duration), nor the integral of time spent in observing all spatial replicates. 
  \Hitem \emph{\textbf{Should actual duration be the total time spent sampling all sites or the amount of time spent sampling per site (e.g. for 5 minute point counts of birds at 10 sites each repeated twice, should we enter 100 minutes (5 minutes X 2 repeats X 10 sites) or 10 minutes (5 minutes X 2 repeats) for duration)?}} As stated above, actual duration is the total spent observing a single point in space, so in this case that would be 10 minutes (then converted to days, so 10 / (60 * 24))).
  \Hitem \emph{\textbf{How do we record duration and actual duration when there are no repeat observations?}} In these cases, study duration is equal to sampling duration. 
   \Hitem \emph{\textbf{How do I record interval in cases where the interval is inconsistent?  For example, in a study were observations were observations were repeated in 1979, 1980, 1981, 1984, 2007, 2009?}} Find the time between each successive period, and then take the average of that (remember to convert to days!). If there are two or more sets of unevenly spaced days for each site/plot/measurement being taken in the study, then find the average interval for each, and average the averages.  
    \Hitem \emph{\textbf{How do you determine the interval for paleo-reconstructions?}} Use the minimum estimate for dating precision as the estimate of time between samples (e.g. 50 years in the study of European charcoal deposits DOI:10.1111/geb.12090). 

@@

    \Hitem \emph{\textbf{What about intervals and durations for instrument-collected, or automatically-collected, observations?}} These are often similar in essence to the paleo-reconstruction case. Take the minimum temperature or daily rainfall recorded at a weather station, which require constant observation over 24 hours to report. In such cases, the interval and sampling duration are both 24 hours. On the other hand, automated logging systems will provide a series of high frequency observations that are collected instantaneously. In these cases the sampling duration should be $\leq$second, and the interval should be the period between successive instantaneous measurements. 
    \Hitem \emph{\textbf{How do you treat interval for a case where repeated samples are taken during a season, across several seasons (e.g.``we performed repeat bird counts every 10 days between March and June of 2005, 2006, and 2008'')?}} Since the sampling is focused on seasons, and presumably some season-specific phenomenon (e.g. breeding behavior), the reported values should be pegged to the season, not averaged across the duration (the start and end dates of the study). So in this example, it would be 10 days.

\hspace{-1cm}\underline{\textbf{Spatial scales:}}
  \Hitem \emph{\textbf{What is resolution, and how do I record it?}} This is the finest scale at which a complete measurement of every unit of the quantity of interest is recorded.  For example, if the measurement in question is a tree stem count, the resolution is determined by the size of the plot used to record every tree stem. Taking this example further, let's say a study reports a plot size of 100 x 100 m, but then goes on to report that they counted stems within a single 1 m wide transect within this larger plot.  In this case, the plot resolution is in fact 100 m x 1 m, or 100 m$^2$ (sampling resolution should be reported in m$^2$).  In another example, if the reported plot size was 20 x 20 m, but the authors in fact only measured a random selection of, say, grass stems on which they counted aphids within those plots, then use an estimate of the area of the grass stem as the sampling resolution \cite{gagic_agricultural_2012}.

  \Hitem \emph{\textbf{What is study extent, and how do I record it?}} Study extent (or, simply, extent) is defined as the total extent bounding all spatial replicates, divided by 10,000 to convert to hectares. For studies in which spatial replicates are not spatially contiguous, this means the area of the minimum polygon bounding all spatial replicates. For example, if the study is conducted in a national park, the effective survey extent would be the area of the national park; if the study is conducted in three national parks, the effective survey extent would be the sum of the areas of all national parks. To calculate the effective survey extent, use the area of the study area/region given in the paper; when the area is not given, but when the survey region is given by name (e.g., Joshua Tree National Park or United States), look up the area through an online search and convert to hectares. When the area is not named, but a map is given, use an appropriate digitizing platform with a suitable map-providing backend (e.g. Google Earth Pro, QGIS with OpenLayers plugin) to navigate to the region and delineate a minimum convex polygon surrounding the plots/transects/sampling units to calculate the area in hectares. When studies focus on portions of survey regions that are clearly distinct from their surroundings (e.g., mangroves in a coastal National Park), try to delineate the focal portions (the mangroves) and not the larger survey region (entire National Park) using Google Earth Pro or a similar application with digitizing and area estimation capabilities.

For spatially contiguous studies (e.g. those based on satellite imagery), the extent is the total area covered by the imagery (in such cases, extent equals actual extent), but only records the area of imagery analyzed by the authors (e.g. if the study area required four Landsat scenes to cover, but covered only the inner quarter of each image, report the extent as the summed area of the four quarters). However, if spatially contiguous studies only use a sub-sample of pixels, extent is the area of the minimum convex polygon bounding all pixels (calculated following the methods above).

@@

  \Hitem \emph{\textbf{What is \emph{actual extent}, and how do I record it?}}  Actual extent is the sampling resolution multiplied by the number of spatial replicates, divided by 10,000 to convert to hectares. For studies in which the spatial replicates are not spatially contiguous (as with most field-based studies), this means resolution (see Q17) multiplied by number of plots. For spatially contiguous studies (e.g., those based on remote sensing imagery), it should be the total area covered by the imagery, i.e., pixel resolution multiplied by the number of pixels. However, as with extent, only record the area analyzed by the authors. If they used a sub-sample of pixels, the actual extent is the number of those pixels multiplied by pixel resolution.

@@

  \Hitem \emph{\textbf{What about studies that sample individual organisms?}} If the study is making a total count of all organisms (let's say a mammal species) within a fixed plot size, or even a variable plot size from which an average plot size (and thus sampling resolution) can be estimated, then follow the procedures described in Q17. However, if the individual animals are the unit of measure (either because a sub-sample of them is being made within a defined plot, or because the observation is not contingent on being located within a plot (maybe a blood sample or body weight is being recorded, for example), then simply estimate two-dimensional area occupied by the animal as the plot resolution, and the number of sampled animals provide the spatial replicates (for calculating actual extent). Occasionally individual animals might be recorded, but within the context of some natural feature, such as a nesting site where the survival of individual chicks is the measurement of interest \cite{roche_relative_2008}. In this case, an estimate of the nest area provides the sampling resolution. In cases where individual animals are tracked using radio or GPS collars, to calculate actual extent, use the number of locational fixes as the quantity of spatial replicates and the animal's two-dimensional body area. If the number of GPS points is not given, the number of fixes can be estimated from the duration during which individuals were collared and the recording interval.

@@

\subsubsection*{Between-observer coefficient of variation}
\vspace{-10pt}
We used the maximum uncertainty values for each dimension from the inter-observer variability analysis (Table S2) to determine the bounds of the random perturbations applied to each record over the 1000 iteration resample.

%\begin{figure}[!ht]
%%\begin{wrapfigure}{c}{1\textwidth}
%\includegraphics[width=1\textwidth]{figures/res_v_extent_pbstrap2.pdf}
%\vspace{-0.15 cm}
%\caption{Two-dimensional kernel density estimates of observational densities within the domains defined by sampling interval and spatial resolution (left) and temporal duration and spatial extent (right). Density estimates were applied to the log-transformed values of each observational dimension after being randomly perturbed within uncertainty bounds drawn from an assessment of between-observer variability. Density estimates are rescaled to represent percentages. }
%\label{afoto1}
%\end{figure}
%

\subsubsection*{Trends in observational methods by year}
\vspace{-10pt}

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\includegraphics[width=1\textwidth]{figures/stype_by_yr.png}
\vspace{10 pt}
\caption{Trends in percentage of observing methods by year of publication. The coefficient of a weighted (by number of studies in each year) linear regression fit to the annual percentages of observations made with remote sensing (left), field methods (center), and automated sensors (right) is presented at the top of each plot, as well as the regression coefficient of determination and p-value. }
\label{afoto1}
\end{figure}

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\center{\includegraphics[width=0.8\textwidth]{figures/dim_by_yr.png}}
\vspace{5 pt}
\caption{Trends in observational scales by year of publication. The coefficient of a weighted (by number of studies in each year) linear regression, fit to the logarithm (base 10) of the mean scale values (calculated for each publication year) for the six assessed dimensions ispresented at the top of each plot, as well as the model coefficient of determination and p-value. }
\label{afoto1}
\end{figure}
%\clearpage

\subsubsection*{Choice of bandwidth in kernel density estimation}
\vspace{-10pt}
Figure S1 indicates the effect that varying bandwidth has on the appearance of the kernel density estimates. The smallest bandwidth (0.4) tested (Fig. S2 top) shows that the primary observational concentrations described in the main text are evident but harder to discern.  Most difficult to discern is the oblong concentration of observations that in Fig. 1A (main text) is bounded on the lower right at monthly to yearly intervals and 100-1000 m$^2$ resolutions and on the upper left by near-daily to monthly observations and 0.1-10 ha resolutions. With the smaller bandwidth this concentration appears as two separate patches (Fig. S2 top left), but with 0.7 bandwidth applied becomes coherent (Fig. S3 middle left).

### 2017-05-24 paper/main.tex
Updated analysis of annual change in observing methods and scales. Revision of main text to accomodate this change, as well as addition of related figures to SI.  Fixes on explanation of methods/FAQ in SI.

\noindent \textbf{\emph{Uncertainties in quantifying observational scales}}\\
This analysis did not necessarily quantify the scales \emph{represented} by ecological observations (SI). Observations that were non-continuous in time or space (e.g. point-based field measurements) may effectively represent larger scales than our estimates suggest, due to phenomenon-dependent factors such as autocorrelation and representativeness of the sampling scheme \cite{underwood_experiments_1997, palmer_scale_1994,cao_comparison_2002, legendre_spatial_1993,collins_method_2000-1}. From a spatial perspective, this concern does not apply to contiguous sampling schemes, which were primarily based on remote sensing and cover the largest area, thus the net effect is that this analysis may underestimate the effective areas of smaller observational extents. From a temporal perspective, our analysis likely underestimates the effective duration of many observations, particularly those where instantaneous, repeated measures of slow-changing ecological features were made. Snapshots in time may be sufficient to capture the temporal dynamics of such phenomena (e.g. changes in vegetation cover \cite{hansen_high-resolution_2013}), and for these the total time elapsed between the first and last measurements (the span of observation) may more closely approximate effective duration. However, many studies focused on more dynamic phenomena, and for these long periods of continuous observation may be more important for understanding dynamics than frequent repeats. For example, wildfire extent and duration can be mapped by daily return satellites \cite{roy_prototyping_2005,jones_fire_2009}, but the instantaneous nature of the imaging means that it cannot be used to directly measure fire behavior \cite{clements_observing_2007}, except patterns resolving themselves at scales greater than 12-24 hours. For this reason, and to provide a consistent standard for estimating duration, we recorded actual duration.

Adding to our motivation to record actual observational scales was the fact that most studies provided insufficient methodological detail to judge whether sampling schemes were suitable matched to and representative of the phenomena they were measuring. The observational scales were not clearly reported in many studies, such that we had to estimate spatial resolution and extent in 62\% and 68\% of cases, respectively, and temporal interval and duration in 40\% and 86\% of cases. Although the confidence intervals around our scale estimates (Fig. 1) suggests that our findings are robust to this level of uncertainty, it would have likely confounded efforts to estimate effective scales.

@@

Despite these caveats, our results provide valuable insight into the spatial and temporal domains being addressed by modern ecological research. Our results show that most observations are collected at small spatial scales, are either unrepeated or relatively infrequent ($\geq$1 month interval), and in aggregate cover relatively narrow periods of time ($\leq$1 month). Very little research is conducted at high spatial and temporal resolutions over large areas or for long time periods, indicating that, despite the well-established understanding of the importance of multi-scale assessments for understanding ecological patterns and processes \cite{levin_problem_1992,wiens_spatial_1989}, efforts focused on larger scales are still relatively sparse within the discipline \cite{levin_problem_1992,wiens_spatial_1989}. Furthermore, the unclear documenting of observational scales implies that scale is not a primary concern in much ecological research \cite{chave_problem_2013, wheatley_factors_2009}. Taken together, this narrowness and poor documentation (a tendency that is also evident in the geographical sciences \cite{margulies_ambiguous_2016}) suggests that the ecological understanding drawn from many of these observations may have limited generalizability \cite{margulies_ambiguous_2016, wheatley_factors_2009, wiens_spatial_1989}, a concern that has been previously noted due to ecology's geographical bias towards anthropogenically undisturbed and temperate ecosystems \cite{martin_mapping_2012}.

>>>>

\noindent \textbf{\emph{Observational methods}}\\
We classified the method used to collect each observation into several broad categories, which were field methods (manual \emph{in situ} data collection), automated (\emph{in situ}) sensing, remote sensing, other geographic data, and paleo-reconstruction approaches. Field methods were used for 80\% of observations, automated sensing for 12.4\%, remote sensing for 6.3\%, and paleo-reconstruction and other geographic data each for less than 1\%. Using linear regression (weighted by the number of observations per publication year) to assess whether the relative frequency of observing methods changed during the 10 year study period, the use of remote sensing appeared to increase by 1.3\% per year from 2004-2014 (R$^2$ = 0.25, p$<$0.12), and field methods declined by the same percentage (R$^2$ = 0.1, p$<$0.18), although both slopes failed to meet the customary threshold for statistical significance. Automated sensing methods showed no trend (SI).

\noindent \textbf{\emph{Potential biases and uncertainties in quantifying observational scales}}\\
%This analysis did not necessarily quantify the scales \emph{represented} by ecological observations (SI). Observations that were non-continuous in time or space (e.g. point-based field measurements) may effectively represent larger scales than our estimates suggest, due to phenomenon-dependent factors such as autocorrelation and representativeness of the sampling scheme \cite{underwood_experiments_1997, palmer_scale_1994,cao_comparison_2002, legendre_spatial_1993,collins_method_2000-1}. From a spatial perspective, this concern does not apply to contiguous sampling schemes, which were primarily based on remote sensing and cover the largest area, thus the net effect is that this analysis may underestimate the effective areas of smaller observational extents. From a temporal perspective, our analysis likely underestimates the effective duration of many observations, particularly those where instantaneous, repeated measures of slow-changing ecological features were made. Snapshots in time may be sufficient to capture the temporal dynamics of such phenomena (e.g. changes in vegetation cover \cite{hansen_high-resolution_2013}), and for these the total time elapsed between the first and last measurements (the span of observation) may more closely approximate effective duration. However, many studies focused on more dynamic phenomena, and for these long periods of continuous observation may be more important for understanding dynamics than frequent repeats. For example, wildfire extent and duration can be mapped by daily return satellites \cite{roy_prototyping_2005,jones_fire_2009}, but the instantaneous nature of the imaging means that it cannot be used to directly measure fire behavior \cite{clements_observing_2007}, except patterns resolving themselves at scales greater than 12-24 hours. For this reason, and to provide a consistent standard for estimating duration, we recorded actual duration.

%Adding to our motivation to record actual observational scales was the fact that 
There were several potential methodological aspects that could have influenced our assessment of ecology's spatial and temporal domains. The first stems from our finding that many studies did not precisely report observational scales, which meant that we had to estimate, rather than simply record, these values for most observations (specifically, in 63, 60, and 67\% of cases for resolution, extent, and actual extent, and 36\%, 64\%, and 83\% of cases for interval, duration, and effective duration). The inevitable errors caused by estimation may have biased our overall findings. However, we attempted to quantify this error by assessing inter-rater disagreement and resampling techniques. The resulting confidence intervals (Fig. 1) suggest that it was unlikely that estimation errors unduly influenced our findings.

Another potential source of bias was the rule set we used to estimate observational scales.

Finally, because our review did not include papers beyond 2014, the omission of studies from the most recent years could have introduced bias into our domain estimates. Indeed, if the trend towards increasing use of remote sensing between 2004-2014 was not spurious, we can project that a repeated study applied to papers published between 2004-2017 would find that remote sensing was used for 7.7\% of observations (a 22\% increase), which would increase the mean effective extent by 17.4\% (95\% CI = -1.3-67\%), or 0.07 orders of magnitude, above that observed here (see SI for details of calculation). Further support for such trend can be found by averaging dimensions by publication year (SI), which reveals that observation extent increased by 0.253 orders of magnitude per year between 2004-2014 (R$^2$ = 0.2.5, p$<$0.07). This somewhat clearer trend also suggests that including more recent studies would show observational extent to be somewhat larger, although in this case by a more modest 5.5\% (0.02 orders of magnitude).

%Additional sources of uncertainty: rules for extracting scales; diversity of study types and ecological methods.

%Most studies provided insufficient methodological detail to judge whether sampling schemes were suitable matched to and representative of the phenomena they were measuring. The observational scales were not clearly reported in many studies, such that we had to estimate spatial resolution and extent in 62\% and 68\% of cases, respectively, and temporal interval and duration in 40\% and 86\% of cases. Although the confidence intervals around our scale estimates (Fig. 1) suggests that our findings are robust to this level of uncertainty, it would have likely confounded efforts to estimate effective scales.

@@

the continued dominance of field-based research, and the limited use of methods that allow larger areas to be comprehensively observed, such as remote sensing.

satellite observations typically have lower information content than field measurements for a given location, and in many cases only provide proxy measures for the ecological features of interest, such as forest understorey structure \cite{estes_remote_2010}, thereby making ecologists less inclined to use the technology \cite{turner_remote_2003}.

Adapt and tone down, retain these points:
Furthermore, the unclear documenting of observational scales implies that scale is not a primary concern in much ecological research \cite{chave_problem_2013, wheatley_factors_2009}. (Scale is not a concern that cuts across the entire discipline of ecology). 
Narrowness and poor documentation (a tendency that is also evident in the geographical sciences \cite{margulies_ambiguous_2016}) 
Ecological understanding drawn from many of these observations may have limited generalizability \cite{margulies_ambiguous_2016, wheatley_factors_2009, wiens_spatial_1989}, 
a concern that has been previously noted due to ecology's geographical bias towards anthropogenically undisturbed and temperate ecosystems \cite{martin_mapping_2012}.

Our results provide valuable insight into the spatial and temporal domains being addressed by modern ecological research. Our results show that most observations are collected at small spatial scales, are either unrepeated or relatively infrequent ($\geq$1 month interval), and in aggregate cover relatively narrow periods of time ($\leq$1 month). Very little research is conducted at high spatial and temporal resolutions over large areas or for long time periods, indicating that, despite the well-established understanding of the importance of multi-scale assessments for understanding ecological patterns and processes \cite{levin_problem_1992,wiens_spatial_1989}, efforts focused on larger scales are still relatively sparse within the discipline \cite{levin_problem_1992,wiens_spatial_1989}. Furthermore, the unclear documenting of observational scales implies that scale is not a primary concern in much ecological research \cite{chave_problem_2013, wheatley_factors_2009}. Taken together, this narrowness and poor documentation (a tendency that is also evident in the geographical sciences \cite{margulies_ambiguous_2016}) suggests that the ecological understanding drawn from many of these observations may have limited generalizability \cite{margulies_ambiguous_2016, wheatley_factors_2009, wiens_spatial_1989}, a concern that has been previously noted due to ecology's geographical bias towards anthropogenically undisturbed and temperate ecosystems \cite{martin_mapping_2012}.
